---
title: "Project on Prediction Of Bank Customer Attrition"
author: "Praveena_Munnam_vxw648
date: "2023-04-08"
output:
  html_document: default
  pdf_document: default
---

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(e1071)
library(partykit)
library(rsample)
library(caret)
library(ROCR)
library(GGally)
library(tidyverse)
library(survival)
library(corrplot)
library(car)
library(MASS)
library(e1071) 
library(ISLR) 
library(rpart)
library(rpart.plot)
library(randomForest)
```


## Reading the dataset
```{r}
bank = read.csv("churn.csv", header = TRUE, stringsAsFactors = T)
```

## Structure of the dataset

```{r}
str(bank)
```
## Checking whether the data set have the null values

```{r}
colSums(is.na(bank))
```

## Data cleaning

```{r}
bank$Exited = as_factor(bank$Exited)
bank$HasCrCard = as_factor(bank$HasCrCard)
bank$IsActiveMember = as_factor(bank$IsActiveMember)
```


## Exploratory data analysis

## Gender Variable

```{r}
ggplot(data = bank) + 
  geom_bar(mapping = aes(x = Exited, fill = Gender), position = "dodge") +
  scale_fill_manual(values = c("palevioletred", "cadetblue"))
```


# Bank products
## HasCrCard
```{r}
ggplot(data = bank,
       aes(x = HasCrCard,
           fill = Exited)) +
  geom_bar() +
  scale_x_discrete(breaks = c(0,1),
                   labels = c("Not Having the Card",
                              "Having the Card"))+
 scale_fill_manual(values = c("palevioletred", "cadetblue"))
```
## IsActiveMember

```{r}
ggplot(data = bank,
       aes(x = IsActiveMember,
           fill = Exited)) +
  geom_bar() +
  scale_x_discrete(breaks = c(0,1),
                   labels = c("Not an Active Member",
                              "Active Member"))+
 scale_fill_manual(values = c("palevioletred", "cadetblue"))
```
## No of products
```{r}
ggplot(data = bank,
       aes(x = NumOfProducts,
           fill = Exited)) +
  geom_bar()+
 scale_fill_manual(values = c("palevioletred", "cadetblue"))
```
## Tenure

```{r}
ggplot(data = bank,
       aes(x = Tenure,
           fill = Exited)) +
  geom_bar()+
 scale_fill_manual(values = c("palevioletred", "cadetblue"))
```

## Continous variable 

```{r}
bank[, names(bank) %in% c('Age', 'Balance', 'CreditScore', 'EstimatedSalary')] %>%
  gather() %>%
  ggplot(aes(value, fill = key)) +
    facet_wrap(~ key, scales = "free") +
    geom_boxplot() +
    theme(axis.text.x = element_text(size = 7, angle=90), axis.text.y = element_text(size = 7))

```


## Exited variable

```{r}
churn <- ggplot(bank, aes(x = factor(ifelse(Exited == 0, "Not Exited", "Exited")))) +
  geom_bar() +
  stat_count(geom = "text", colour = "pink", aes(label = paste("N =", after_stat(count))), 
             position = position_stack(vjust = 0.5)) +
  xlab("Choice") + ylab("Count")
churn
```
## Exited in percentage
```{r}
churn1 <- bank %>% 
  group_by(Exited) %>% 
  summarize(count = n()) %>% 
  mutate(pct = count/sum(count))

ggplot(churn1, aes(x = factor(Exited), y = pct, fill = factor(Exited))) + 
  geom_bar(stat = "identity", show.legend = FALSE) + 
  geom_text(aes(label = scales::percent(pct)), position = position_stack(vjust = 0.5)) +
  scale_y_continuous(labels = scales::percent) +
 scale_fill_manual(values = c("palevioletred", "cadetblue")) +
  ggtitle("Percentage of Customers Who Exited and Not Exited their Accounts") +
  xlab("Exited") +
  ylab("Percent of Customers") + 
  theme_minimal()
```

## Check for correlation
```{r}
bank_num = select_if(bank, is.numeric)
C = cor(bank_num)
```



# Plot the correlation matrix
```{r}
corrplot(C, method = "number")
```

## Selecting the variables

### Removing variables which are irrelavant

```{r}
bank = dplyr::select(bank,-RowNumber)
bank = dplyr::select(bank,-CustomerId)
bank = dplyr::select(bank,-Surname)
```


```{r}
names(bank)
```

```{r}
bank$Exited <- as_factor(make.names(as.character(bank$Exited)))
bank$HasCrCard <- as_factor(make.names(as.character(bank$HasCrCard)))
bank$IsActiveMember <- as_factor(make.names(as.character(bank$IsActiveMember)))
```

```{r}
levels(bank$HasCrCard)
```
## Training and Testing

```{r}
set.seed(444)
split <- sample(nrow(bank), nrow(bank)*0.8)
banktrain <- bank[split,]
banktest <- bank[-split,]
```


```{r}
table(banktrain$Exited)
```

### DOWNSAMPLE

```{r}
traindown <- downSample(x = subset(banktrain, select = -Exited),
                        y = banktrain$Exited,
                        list = F,
                        yname = "Exited")

table(traindown$Exited)
```
```{r}
levels(banktrain$Exited)
```


## Building the models


```{r}
set.seed(100)
ctrl <- trainControl(method = "LGOCV",
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = TRUE)
```

## Logistic Regression

```{r}
 set.seed(4)
 logisticTune <-train(Exited ~ ., 
                 data = traindown,
 method = "glm",
 metric = "ROC",
 trControl = ctrl)
 logisticTune 
```

```{r}
pred_log <- predict(logisticTune, newdata= banktest, type = "prob")[, 1]
confusionmatrix.log = confusionMatrix(data = predict(logisticTune, banktest), reference = banktest$Exited)
confusionmatrix.log
```

```{r}
set.seed(15)
logpred <- prediction(pred_log, banktest$Exited)
logperf <- performance(logpred, "tpr", "fpr")
plot(logperf, colorize = TRUE)
abline(a = 0, b = 1)
auc_log <- round(as.numeric(performance(logpred, "auc")@y.values), 2)
legend(0.8, 0.2, paste0("AUC = ", auc_log),  col = "red", cex = 1)
```



## Flexible Discriminate Analysis

```{r}
 set.seed(476)
 FDATune <- train(Exited ~ ., 
                 data = traindown,
 method = "fda",
 preProc = c('center', 'scale'),
 metric = "ROC",
 trControl = ctrl)
 FDATune
```
```{r}
plot(FDATune)
```

```{r}
predclass_FDA = predict(FDATune, newdata = banktest, type = "prob")[,1]
confusionmatrix.fda = confusionMatrix(data = predict(FDATune, banktest), reference = banktest$Exited)
confusionmatrix.fda
```

```{r}
Pred_val3 <- prediction(predclass_FDA, banktest$Exited)
ROCperf3 <- performance(Pred_val3, "tpr", "fpr")
plot(ROCperf3, colorize = TRUE, col = "red")
abline(0, 1, lty = 2)
auc_fda <- round(as.numeric(performance(Pred_val3, "auc")@y.values[[1]]), 2)
text(0.8, 0.2, paste0("AUC = ", auc_fda), col = "black", cex = 1)
```

## Penalized Models
```{r}
glmnGrid <- expand.grid(alpha = c(0, .1, .2, .4, .6, .8, 1),
                        lambda = seq(.01, .2, length = 40))
set.seed(476)
glmnTune = train(Exited ~ ., 
                 data = traindown,
                 method = "glmnet",
                 tuneGrid = glmnGrid,
                 metric = "ROC",
                 trControl = ctrl)
```

```{r}
plot(glmnTune)
```



```{r}
pred_glmn <- predict(glmnTune, newdata= banktest, type = "prob")[, 1]
confusionmatrix.glmn = confusionMatrix(data = predict(glmnTune, banktest), reference = banktest$Exited)
confusionmatrix.glmn
```

```{r}
set.seed(1)
Pred_val1 <- prediction(pred_glmn, banktest$Exited) 
ROCperf1 <- performance(Pred_val1, "tpr", "fpr")
plot(ROCperf1, colorize = TRUE)
abline(0, 1, lty = 2)
auc_glmn <- round(as.numeric(performance(Pred_val1, "auc")@y.values[[1]]), 2)
text(0.8, 0.2, paste0("AUC = ", auc_glmn), col = "black", cex = 1)
```

## KNN model

```{r}
library(caret)
set.seed(123)
modelknn <- train(Exited ~ ., 
                  data = traindown, 
                  method = "knn",
                  preProcess = c("center", "scale"), 
                  tuneLength = 10,
                  trControl = ctrl)
print(modelknn)
```

```{r}
predknn <- predict(modelknn, newdata = banktest, type = "raw")
confusionmatrix.knn <- confusionMatrix(predknn, banktest$Exited)
confusionmatrix.knn
```

```{r}
set.seed(1)
pred_knn <- predict(modelknn, newdata= banktest, type = "prob")[, 1]
Pred_val5 <- prediction(pred_knn, banktest$Exited) 
ROCperf5 <- performance(Pred_val5, "tpr", "fpr")
plot(ROCperf5, colorize = TRUE, xlab = "False Positive Rate", ylab = "True Positive Rate", main = "ROC Curve for KNN Model")
abline(0, 1, lty = 2)
auc_knn <- round(as.numeric(performance(Pred_val5, "auc")@y.values[[1]]), 2)
text(0.8, 0.2, paste0("AUC = ", auc_knn), col = "Black", cex = 1)
```


## Naïve Bayes

```{r, include=FALSE}
 set.seed(476)
 NBTune <- train(Exited ~ ., 
                 data = traindown,
 method = "nb",
 preProc = c('center', 'scale'),
 metric = "ROC",
 trControl = ctrl)
```
```{r}
NBTune
```

```{r}
pred_NB <- predict(NBTune, newdata= banktest, type = "prob")[, 1]
confusionmatrix.NB = confusionMatrix(data = predict(NBTune, banktest), reference = banktest$Exited)
confusionmatrix.NB
```

```{r}
set.seed(1)
Pred_val8 <- prediction(pred_NB, banktest$Exited) 
ROCperf8 <- performance(Pred_val8, "tpr", "fpr")
plot(ROCperf8, colorize = TRUE)
abline(0, 1, lty = 2)
auc_NB <- round(as.numeric(performance(Pred_val8, "auc")@y.values[[1]]), 2)
text(0.8, 0.2, paste0("AUC = ", auc_NB), col = "black", cex = 1)
```

## Bagging 


```{r}
set.seed(476)
treebagTune = train(Exited ~ ., 
                data = traindown,
                     method = "treebag",
                     nbagg = 50,
                     trControl = ctrl)

treebagTune
```
```{r}
pred_bg <- predict(treebagTune, newdata= banktest, type = "prob")[, 1]
confusionmatrix.bg = confusionMatrix(data = predict(treebagTune, banktest), reference = banktest$Exited)
confusionmatrix.bg
```

```{r}
set.seed(1)
Pred_val11 <- prediction(pred_bg, banktest$Exited) 
ROCperf11 <- performance(Pred_val11, "tpr", "fpr")
plot(ROCperf11, colorize = TRUE)
abline(0, 1, lty = 2)
auc_bg <- round(as.numeric(performance(Pred_val11, "auc")@y.values[[1]]), 2)
text(0.8, 0.2, paste0("AUC = ", auc_bg), col = "black", cex = 1)
```


## Random forest
```{r}
mtryGrid <- data.frame(mtry = 1:10) #since we only have 10 predictors

### Tune the model using cross-validation
set.seed(476)
rfTune <- train(Exited ~ ., 
                data = traindown,
                method = "rf",
                tuneGrid = mtryGrid,
                ntree = 200,
                importance = TRUE,
                trControl = ctrl)
rfTune
```

```{r}
pred_rf <- predict(rfTune, newdata= banktest, type = "prob")[, 1]
confusionmatrix.rf1 = confusionMatrix(data = predict(rfTune, banktest), reference = banktest$Exited)
confusionmatrix.rf1
```

```{r}
set.seed(1)
Pred_val9 <- prediction(pred_rf, banktest$Exited) 
ROCperf9 <- performance(Pred_val9, "tpr", "fpr")
plot(ROCperf9, colorize = TRUE)
abline(0, 1, lty = 2)
auc_rf1 <- round(as.numeric(performance(Pred_val9, "auc")@y.values[[1]]), 2)
text(0.8, 0.2, paste0("AUC = ", auc_rf1), col = "black", cex = 1)
```


```{r}
varImp(rfTune)
```

```{r}
rf_object <- rfTune$finalModel
var_imp <- importance(rf_object, type = 2)
var_imp_df <- data.frame(Variables = row.names(var_imp),
                         Importance = round(var_imp[, "MeanDecreaseGini"], 2)) %>%
              arrange(desc(Importance))
ggplot(var_imp_df, aes(x = reorder(Variables, Importance), y = Importance)) + 
  geom_bar(stat = "identity", fill = "cadetblue") +
  coord_flip() +
  labs(title = "Variable Importance Plot", y = "Importance", x = "Predictor Variable")
``` 


## Boosting

```{r}
gbmGrid = expand.grid( interaction.depth = seq( 1, 7, by=2 ),
                       n.trees = seq( 100, 1000, by=100 ),
                       shrinkage = c(0.01, 0.1),
                       n.minobsinnode = 10 )

set.seed(476) 
gbmTune <- train(Exited ~ ., 
                data = traindown,
                 method = "gbm",
                 tuneGrid = gbmGrid,
                 trControl = ctrl,
                 verbose = FALSE)
```
```{r}
plot(gbmTune, auto.key = list(columns = 4, lines = TRUE))
```

```{r}
pred_bo <- predict(gbmTune, newdata= banktest, type = "prob")[, 1]
confusionmatrix.bo = confusionMatrix(data = predict(gbmTune, banktest), reference = banktest$Exited)
confusionmatrix.bo
```

```{r}
set.seed(1)
Pred_val10 <- prediction(pred_bo, banktest$Exited) 
ROCperf10 <- performance(Pred_val10, "tpr", "fpr")
plot(ROCperf10, colorize = TRUE)
abline(0, 1, lty = 2)
auc_bo <- round(as.numeric(performance(Pred_val10, "auc")@y.values[[1]]), 2)
text(0.8, 0.2, paste0("AUC = ", auc_bo), col = "black", cex = 1)
```


```{r}
pred1.accuracy =confusionmatrix.log$overall[1]
pred3.accuracy = confusionmatrix.fda$overall[1]
pred4.accuracy = confusionmatrix.glmn$overall[1]
pred5.accuracy = confusionmatrix.knn$overall[1]
pred9.accuracy = confusionmatrix.NB$overall[1]
pred12.accuracy = confusionmatrix.bg$overall[1]
pred10.accuracy = confusionmatrix.rf1$overall[1]
pred11.accuracy = confusionmatrix.bo$overall[1]
```



## Model Evaluation
```{r}
results <- data.frame(
  Model = c("Logistic Regression", "Flexible discriminant analysis", "Penalized Models", "K-nearest neighbors", "Naive Bayes","Bagging", "Random Forest", "Boosting"),
  Accuracy = sprintf("%.2f", c(pred1.accuracy, pred3.accuracy, 
                               pred4.accuracy, pred5.accuracy,pred9.accuracy,pred12.accuracy,                          pred10.accuracy,pred11.accuracy)),
  Sensitivity = sprintf("%.2f", c(confusionmatrix.log$byClass["Sensitivity"],
                                   confusionmatrix.fda$byClass["Sensitivity"],
                                   confusionmatrix.glmn$byClass["Sensitivity"],
                                   confusionmatrix.knn$byClass["Sensitivity"],
                                  confusionmatrix.NB$byClass["Sensitivity"],
                                  confusionmatrix.bg$byClass["Sensitivity"],
                                  confusionmatrix.rf1$byClass["Sensitivity"],
                                  confusionmatrix.bo$byClass["Sensitivity"])),
  Specificity = sprintf("%.2f", c(confusionmatrix.log$byClass["Specificity"],
                                   confusionmatrix.fda$byClass["Specificity"],
                                   confusionmatrix.glmn$byClass["Specificity"],
                                   confusionmatrix.knn$byClass["Specificity"],
                                  confusionmatrix.NB$byClass["Specificity"],
                                  confusionmatrix.bg$byClass["Specificity"],
                                  confusionmatrix.rf1$byClass["Specificity"],
                                  confusionmatrix.bo$byClass["Specificity"])),
  AUC = sprintf("%.2f", c(as.numeric(performance(logpred, "auc")@y.values),
                           as.numeric(performance(Pred_val3, "auc")@y.values[[1]]),
                           as.numeric(performance(Pred_val1, "auc")@y.values[[1]]), 
                           as.numeric(performance(Pred_val5, "auc")@y.values[[1]]),
                          as.numeric(performance(Pred_val8, "auc")@y.values[[1]]),
                          as.numeric(performance(Pred_val11, "auc")@y.values[[1]]),
                          as.numeric(performance(Pred_val9, "auc")@y.values[[1]]),
                          as.numeric(performance(Pred_val10, "auc")@y.values[[1]])))
)

results

```





#Resamples of Tranining data

```{r}
res = resamples(list(logistic = logisticTune, FDA = FDATune , Penalized = glmnTune , knn = modelknn,  NaiveBayes = NBTune, RandomForest = rfTune , Boosting = gbmTune, Bagging = treebagTune))
dotplot(res)
```





